Jane, John … Leslie? A Historical Method for Algorithmic Gender Prediction

### 2014-09-22 article.Rmd
Minor edits; conclusion

One 2013 analysis of 2,500 recent history Ph.D.'s found that "gender played little role in employment patterns across particular professions and industries."^[http://historians.org/Documents/Many_Careers_of_History_PhDs_Final.pdf] Our own analysis reveals a more complicated reality. The flagship journal of professional historians continues to publish twice as many reviews of books by female authors as male authors. This disparity is put into even sharper relief when set against the relative parity achieved by women in producing history dissertations. Women might be getting hired at comparable rates to men, but discriminatory gatekeeping remains alive and well in the historical profession. Our program allows us to quantitatively measure the scope of the problem on a large scale, how it's changed over time, and what remains to be done.

@@

In conclusion, our algorithm for predicting gender from first names combined with historical datasets in the Gender package provides a powerful new method for historians interested in the study of gender. The digital turn has made available very large datasets, yet in the midst of abundance historians always have to deal with the problem of scarcity in the sources. Our methods uses the abundant datasets to produce more information from scarce datasets. The temporal dimension of our work matches a historical question to a historical method, providing much improved over simple, anachronistic lookup methods. This method is extensible to any time or place for which there is a suitable dataset of names from which to make the predictions. As we have demonstrated in our study of gatekeeping in the American historical profession, our method can reveal important details about gender even from datasets which do not otherwise include that information, and it can do so over long historical periods.

>>>>

One 2013 analysis of 2,500 recent history Ph.D.'s found that "gender played little role in employment patterns across particular professions and industries."^[http://historians.org/Documents/Many_Careers_of_History_PhDs_Final.pdf] Our own analysis reveals a more complicated reality. The flagship journal of professional historians continues to publish twice as many reviews of books by male authors as female authors. This disparity is put into even sharper relief when set against the relative parity achieved by women in producing history dissertations. Women might be getting hired at comparable rates to men, but discriminatory gatekeeping remains alive and well in the profession. Our program allows us to quantitatively measure the scope of the problem on a large scale, how it's changed over time, and what remains to be done.

@@

Our algorithm for predicting gender from first names, when combined with historical datasets in the Gender package, provides a powerful new method for researchers interested in the study of gender. The digital turn has made available very large datasets, yet in the midst of abundance many humanists continue to deal with the problem of scarcity in their sources. Our method uses the abundant existing datasets to infer more information from scarce datasets. The temporal dimension of our work matches historical questions to a historical method, providing much improved results over simple, anachronistic lookup methods. This method is extensible to any time or place for which there is a suitable dataset of names from which to make predictions about gender and naming practices. As we have demonstrated in our study of gatekeeping in the American historical profession, our method can reveal important details about gender even from datasets which do not otherwise include that information, and it can do so over long historical periods.
